Modeling the impact of IT- and AI-related emissions in a university setting

### 2025-08-04 doc/paper.tex
Results checkpoint

\documentclass[11pt]{article}
\usepackage[margin=0.9in]{geometry}
\usepackage[T1]{fontenc}
\usepackage{libertine}
\renewcommand{\sfdefault}{fvs} % Use as sans-serif
\renewcommand{\ttdefault}{fvm} % Use as monospace
\renewcommand{\ttfamily}{\small\fontfamily{\ttdefault}\selectfont} % Slightly smaller monospace font
\newcommand{\eg}{{\em e.g.}}
\newcommand{\coe}{CO$_2$e}
\newcommand{\gcoe}{g \coe}
\newcommand{\kgcoe}{k\gcoe}
\newcommand{\gcoekwh}{\gcoe\ / kWh}
%\usepackage{helvet} % Helvetica font for sans-serif
%\usepackage{courier} % Courier font for monospace
\usepackage{graphicx}
\usepackage{hyperref}
\hypersetup{
    hidelinks
}
\usepackage[style=numeric]{biblatex}
\usepackage{titlesec}
\usepackage{enumitem}
\usepackage{titling}

@@

Information technology accounts for over 5\% of global electricity consumption and at least 1.7\% of global carbon emissions~\cite{wb:itu:ict}. At the University of Zürich (UZH), which with over 38,000 students and staff is the largest Swiss university, IT in 2024 accounted for about 1.5 Mt \coe\ (megatons of carbon dioxide equivalent) of greenhouse gas (GHG) emissions, approximately 6.3\% of total university emissions in 2024~\cite{uzh:sustainability:report}. The cited report takes into account university-owned workstations, servers, and audiovisual equipment, but does include other IT-related activities such as networking, cloud computing, or the growing use of artificial intelligence, and in particular the adoption of generative AI and large language models (LLMs).

@@

Using only the types of devices reported in the UZH 2024 Sustainability Report~\cite{uzh:sustainability:report}, and restricting embodied emissions to the Ecoinvent~\cite{ecoinvent} model used by that report, our model estimates emissions of 1,460 t \coe, within 1\% of the 1,454 estimated in the UZH report, even though the methodology is somewhat different. Our model exposes average device power ratings and duty cycle, grid emissions intensity, etc., directly, whereas the approach used in the sustainability report relies on gross emissions factors (\kgcoe\ / device for embodied emissions, \kgcoe\ / device / day for operational emissions) provided by Ecoinvent.

@@

\begin{figure}[h]
\centering
\includegraphics[width=0.8\textwidth]{fig1.png}
\caption{Baseline results}
\label{fig:1}
\end{figure}

@@

@@

  \item Under the assumption that the vast majority (90\%) of UZH students and staff have a laptop and smartphone, manufacturing personal devices accounts for over half of UZH's IT carbon footprint. For the estimated 32,000 laptops and smartphones, embodied emissions total $\sim 2400$ t \coe\ annually. Usage-related emissions amount to about a tenth of that.

@@

  \item Embodied emissions are an important consideration for university-owned devices as well. If we exclude personal devices and LLMs, then total IT emissions shrink to $\sim 1500$ t \coe, of which $\frac{2}{3}$ comprises the embodied emissions of classroom computers, wall displays, printers, etc.

  \item Data center cooling and efficiency represents a potential savings opportunity. The largest operational emissions are from data center servers and cooling (320 t). UZH data centers currently require as much energy for cooling as for computation, compared to the $6-15\%$ cooling overhead of cloud industry leaders. Moving workloads to efficient cloud providers could reduce data center emissions from $\sim 560$ t to $\sim 135$ t \coe\ annually, though this estimate requires further validation.

>>>>

\documentclass[11pt]{article}
\usepackage[margin=0.9in]{geometry}
\usepackage[T1]{fontenc}
\usepackage{libertine}
\renewcommand{\sfdefault}{fvs} % Use as sans-serif
\renewcommand{\ttdefault}{fvm} % Use as monospace
\renewcommand{\ttfamily}{\small\fontfamily{\ttdefault}\selectfont} % Slightly smaller monospace font
\newcommand{\eg}{{\em e.g.}}
\newcommand{\coe}{CO$_2$e}
\newcommand{\gcoe}{g \coe}
\newcommand{\tcoe}{t \coe}
\newcommand{\kgcoe}{k\gcoe}
\newcommand{\gcoekwh}{\gcoe\ / kWh}
%\usepackage{helvet} % Helvetica font for sans-serif
%\usepackage{courier} % Courier font for monospace
\usepackage{graphicx}
\usepackage{hyperref}
\hypersetup{
    hidelinks
}
\usepackage[style=numeric]{biblatex}
\usepackage{titlesec}
\usepackage{enumitem}
\usepackage{titling}

@@

Information technology accounts for over 5\% of global electricity consumption and at least 1.7\% of global carbon emissions~\cite{wb:itu:ict}. At the University of Zürich (UZH), which with over 38,000 students and staff is the largest Swiss university, IT in 2024 accounted for about 1.5 M\tcoe\ (megatons of carbon dioxide equivalent) of greenhouse gas (GHG) emissions, approximately 6.3\% of total university emissions in 2024~\cite{uzh:sustainability:report}. The cited report takes into account university-owned workstations, servers, and audiovisual equipment, but does include other IT-related activities such as networking, cloud computing, or the growing use of artificial intelligence, and in particular the adoption of generative AI and large language models (LLMs).

@@

Using only the types of devices reported in the UZH 2024 Sustainability Report~\cite{uzh:sustainability:report}, and restricting embodied emissions to the Ecoinvent~\cite{ecoinvent} model used by that report, our model estimates emissions of 1,460 \tcoe, within 1\% of the 1,454 estimated in the UZH report, even though the methodology is somewhat different. Our model exposes average device power ratings and duty cycle, grid emissions intensity, etc., directly, whereas the approach used in the sustainability report relies on gross emissions factors (\kgcoe\ / device for embodied emissions, \kgcoe\ / device / day for operational emissions) provided by Ecoinvent.

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=0.8\textwidth]{fig-baseline.png}
  \caption{Baseline results}
  \label{fig:1}
\end{figure}

Many studies~\cite{wb:itu:ict,umich:factsheet} agree that a large fraction of computer and mobile phone energy demand stems from the manufacturing phase. This is due in part to the fact that such end-user devices are not used continuously. Our estimate of embodied emissions comprising approximately 85\% of total laptop emissions is near the top of the range in the literature, and it is consistent with the fact that the carbon intensity of Swiss electricity is about one quarter the EU average~\cite{electricitymaps}.

@@

Hypothetically moving all UZH server workloads from on-premise data centers to a commercial cloud provider eliminates about 170 \tcoe\ of cooling overhead, and replaces 415 \tcoe\ of total server emissions with 380 \tcoe\ of cloud emissions.

\begin{figure}[h]
  \centering
  \includegraphics[width=0.8\textwidth]{fig-cloud.png}
  \caption{Migration to cloud}
  \label{fig:1}
\end{figure}

For relatively small and inefficiently cooled facilities such as those of UZH, migration to the cloud certainly improves power usage effectiveness. Whether the migration is a net win, however, depends on details that would require an in-depth engagement with the cloud provider. The results above assume a one-for-one swap of on-premise servers with cloud ones, as well as the same electricity emissions intensity for the two facilities. If the cloud footprint needed to be bigger, or if the data center were located in a region with dirtier electricity, those factors might outweight the improved cooling efficiency.

There is also a financial argument to consider: if migration to the cloud is more expensive than continued on-premise operation, the savings from not migrating might be invested in more cost effective emissions improvements (including ones not related to IT).

@@

  \item Under the assumption that the vast majority (90\%) of UZH students and staff have a laptop and smartphone, manufacturing personal devices accounts for over half of UZH's IT carbon footprint. For the estimated 32,000 laptops and smartphones, embodied emissions total $\sim 2400$ \tcoe\ annually. Usage-related emissions amount to about a tenth of that.

@@

  \item Embodied emissions are an important consideration for university-owned devices as well. If we exclude personal devices and LLMs, then total IT emissions shrink to $\sim 1500$ \tcoe, of which $\frac{2}{3}$ comprises the embodied emissions of classroom computers, wall displays, printers, etc.

  \item Data center cooling and efficiency represents a potential savings opportunity. The largest operational emissions are from data center servers and cooling (320 t). UZH data centers currently require as much energy for cooling as for computation, compared to the $6-15\%$ cooling overhead of cloud industry leaders. Moving workloads to efficient cloud providers could reduce data center emissions from $\sim 560$ t to $\sim 135$ \tcoe\ annually, though this estimate requires further validation.
